Concurrent Programming in Harmony

### 2021-02-01 book/paper.tex
Pass over Chapters 1-10

In \autoref{fig:tas} we have shown a solution based on a shared
variable and a private variable for each thread.   The private
variables themselves are actually implemented as shared variables,
but they are accessed only by their respective threads.
A thread usually does not keep explicit track of whether it has a lock
or not because it is usually implied by the control flow of the program.
So there is no need to keep \textit{private} as a shared
variable---we only did so to be able to show and check the invariants.
\autoref{fig:tas2} shows a more straightforward implementation of spinlock.
The solution is similar to the na\"{i}ve solution of \autoref{fig:uplock},
but uses test-and-set to check and set the lock variable atomically.
This approach is general for any number of threads.

@@

In Harmony, \texttt{atomic} statements allow you to \emph{implement} your own
interlock primitives such as test-and-set.  Other common examples
include compare-and swap and fetch-and-add.  \texttt{atomic} statements
are not intended to \emph{replace} locks or other synchonization primitives.
When solving synchronization problems you should not directly use
\texttt{atomic} statements but use the synchronization primitives that are available
to you.  But if you want to design a new synchronization primitive, then
use \textbf{atomic} by all means.
You can also use \textbf{atomic} statements in your test code.
In fact, as mentioned before, \texttt{assert} statements are executed atomically.

@@

Harmony has a module called \texttt{synch} that includes support for
binary semaphores and locks.
\autoref{fig:spinlocks} shows how they are implemented, and
\autoref{fig:incfixed} gives an example of how they may be used,
in this case to fix the program of \autoref{fig:inc}.
Notice that the module hides the implementation of
binary semaphores.
The \texttt{synch} module includes a variety of other useful
synchronization primitives, which will be discussed in later
chapters.

@@

In most operating systems, threads are virtual (as opposed to ``raw CPU cores'')
and can be suspended until some condition changes.
For example, a thread that is trying to acquire a lock can be suspended until the lock is
available.
In Harmony, a thread can suspend itself and save its context (state) in a
shared variable.  Recall that the context of a thread contains
its program counter, stack, and registers.
A context is a regular (if complex) Harmony value.
The syntax of the expression is as follows:

@@

There is a second version of the \texttt{synch} module, called \texttt{synchS},
that uses suspension instead of spinlocks.
\autoref{fig:suspension} shows the same interface implemented
using suspension.
The interface is implemented as follows:
\begin{itemize}
\item A binary semaphore maintains both a boolean indicating whether the
binary semaphore is currently acquired and a list of contexts of threads that want to
acquire the binary semaphore.
\item
\texttt{acquire}()
\index{acquire}%
acquires the binary semaphore if available and suspends the invoking thread if not.
In the latter case, the context of the thread is added to the end of the list of contexts.
Note that \textbf{stop} is called within an \textbf{atomic} statement---this is
the only exception to an atomic statement running to completion.  While the
thread is running no other threads can run, but when the thread suspends itself
other threads can run.
\item
\texttt{release}()
\index{release}%
checks to see if any threads are waiting to acquire the binary semaphore.
If so, it uses the \texttt{head} and \texttt{tail}
methods from the \texttt{list} module (see \autoref{ap:module})
to resume the first thread that got
suspended and to remove its context from the list.
\end{itemize}
Selecting the first thread is a design choice.  Another implementation could
have picked the last one, and yet another implementation could have used
\texttt{choose} to pick an arbitrary one.  Selecting the first is a common
choice in binary semaphore implementations as it prevents \emph{starvation}:
\index{starvation}%
every thread
gets a chance to acquire the binary semaphore (assuming every thread eventually releases
it).  \autoref{ch:starvation} will talk more about starvation and how
to prevent it.

@@

\section*{Exercises}
\begin{problems}
\item
Run \autoref{fig:incfixed} using (i) \texttt{synch} and (ii) \texttt{synchS}.
Report how many states were explored by Harmony for each module.
\item \label{ex:xy} \autoref{fig:xy} shows a Harmony program with two variables $x$
(initially 0) and $y$ (initially 100) that can be accessed through methods
\texttt{setX} and \texttt{getXY}.  An application invariant is that \texttt{getXY}
should return a pair that sums to 100.  Add the necessary synchronization code.
\item \label{ex:trylock} Implement \texttt{tryAcquire}(\textit{b}) as an additional
interface for both the \texttt{synch} and \texttt{synchS} modules.
This interface is like \texttt{acquire}(\textit{b}) but never blocks.  It
returns \texttt{True} if the binary semaphore was available (and now acquired) or \texttt{False}
if the binary semaphore was already acquired.
Hint: you do not have to change the existing code.
\item People who use an ATM often first check their balance and then withdraw
a certain amount of money not exceeding their balance.  A negative balance
is not allowed.  \autoref{fig:atm} shows two operations on bank accounts:
one to check the balance and one to withdraw money.
Note that all operations on accounts are carefully protected by a lock
(i.e., there are no data races).
The \texttt{customer}
method models going to a particular ATM and withdrawing money not exceeding
the balance.  Method \texttt{checker} looks for negative balances.
Running the code through Harmony reveals that there is a bug.
It is a common type of concurrency bug known as \emph{Time Of Check Time Of
Execution} (TOCTOE).
\index{Time Of Check Time Of Execution}
\index{TOCTOE}
In this case, by the time the withdraw operation is performed,
the balance can have changed.
Fix the code in \autoref{fig:atm}.  Note, you should leave the
customer code the same. You are only allowed to change the
\texttt{atm\_} methods, and you cannot use the \textbf{atomic}
statement.
\end{problems}

@@

We will first implement the queue as a linked list.
The implementation in \autoref{fig:queue}
uses the \texttt{alloc} module for dynamic allocation
\index{dynamic allocation}%
of nodes in the list using \texttt{malloc}() and \texttt{free}().
\texttt{malloc}($v$) returns a new memory location initialized to $v$,
which should be released with \texttt{free}() when it is no longer in use.
The queue maintains a \texttt{head} pointer to the first element in the list
and a \texttt{tail} pointer to the last element in the list.
The \texttt{head} pointer is \textbf{None} if and only if the queue is empty.
(\textbf{None} is a special address value that is not the address of any
memory location.)

\texttt{queue.Queue()} returns a new queue object consisting of a \textbf{None} head
and tail pointer and a lock.
$\mathtt{queue.put}(q, v)$ and $\mathtt{queue.get}(q)$ both take a pointer $q$ to the
queue object because both may modify the queue.
Before they access the value of the head or tail of the queue they first obtain
the lock.
When they are done, they release the lock.

The implementation in \autoref{fig:queueMS}~\cite{MS98} is similar but uses
separate locks for the head and tail, allowing a \texttt{put} and a \texttt{get} operation
to proceed concurrently.  To avoid contention between the head and the tail,
the queue uses a dummy node at the head of the linked list.
Except initially, the dummy node is the last node that was dequeued.
Note that neither the \texttt{head} nor \texttt{tail} pointer are
ever \textbf{None}.

>>>>

In \autoref{fig:tas} we have shown a solution based on a shared
variable and a private variable for each thread.   The private
variables themselves are actually implemented as shared variables,
but they are accessed only by their respective threads.
A thread usually does not keep explicit track of whether it has a lock
or not because it is usually implied by the control flow of the program---a
thread implicitly \emph{knows} that when it is executing in a critical
section it has the lock.
So there is no need to keep \textit{private} as a shared
variable---we only did so to be able to show and check the invariants.
\autoref{fig:tas2} shows a more straightforward implementation of spinlock.
The solution is similar to the na\"{i}ve solution of \autoref{fig:uplock},
but uses test-and-set to check and set the lock variable atomically.
This approach is general for any number of threads.

@@

In Harmony, \texttt{atomic} statements allow you to \emph{implement} your own
interlock primitives such as test-and-set.  Other common examples
include compare-and swap and fetch-and-add.  \texttt{atomic} statements
are not intended to replace locks or other synchonization primitives.
When solving synchronization problems you should not directly use
\texttt{atomic} statements but use the synchronization primitives that are available
to you.  But if you want to design a new synchronization primitive, then
use \textbf{atomic} by all means.
You can also use \textbf{atomic} statements in your test code.
In fact, as mentioned before, \texttt{assert} statements are executed atomically.

@@

Harmony has a module called \texttt{synch} that includes support for
binary semaphores and locks.
\autoref{fig:spinlocks} shows how they are implemented, and
\autoref{fig:incfixed} gives an example of how they may be used,
in this case to fix the program of \autoref{fig:inc}.
Notice that the module hides the implementation of
binary semaphores and locks.
The \texttt{synch} module includes a variety of other useful
synchronization primitives, which will be discussed in later
chapters.

@@

In most operating systems, threads are virtual (as opposed to ``raw CPU cores'')
and can be suspended until some condition changes.
For example, a thread that is trying to acquire a lock can be suspended until the lock is
available.
In Harmony, a thread can suspend itself and save its context (state) in a
shared variable.  Recall that the context of a thread contains
its program counter, stack, and registers.
A context is a regular (if complex) Harmony value.
The syntax of the expression that a thread executes to suspend itself
is as follows:

@@

There is a second version of the \texttt{synch} module, called \texttt{synchS},
that uses suspension instead of spinlocks.
\autoref{fig:suspension} shows the same interface
as \autoref{fig:spinlocks} implemented using suspension.
The interface is implemented as follows:
\begin{itemize}
\item A binary semaphore maintains both a boolean indicating whether the
binary semaphore is currently acquired and a list of contexts of threads that want to
acquire the binary semaphore.
\item
\texttt{acquire}()
\index{acquire}%
acquires the binary semaphore if available and suspends the invoking thread if not.
In the latter case, the context of the thread is added to the end of the list of contexts.
Note that \textbf{stop} is called within an \textbf{atomic} statement---this is
the only exception to an atomic statement running to completion.  While the
thread is running no other threads can run, but when the thread suspends itself
other threads can run.
\item
\texttt{release}()
\index{release}%
checks to see if any threads are waiting to acquire the binary semaphore.
If so, it uses the \texttt{head} and \texttt{tail}
methods from the \texttt{list} module (see \autoref{ap:module})
to resume the first thread that got
suspended and to remove its context from the list.
\end{itemize}
Selecting the first thread is a design choice.  Another implementation could
have picked the last one, and yet another implementation could have used
\texttt{choose} to pick an arbitrary one.  Selecting the first is a common
choice in binary semaphore implementations as it prevents \emph{starvation}:
\index{starvation}%
every thread
gets a chance to acquire the binary semaphore (assuming every thread eventually releases
it).  \autoref{ch:starvation} will talk more about starvation and how
to prevent it.

@@

\section*{Exercises}
\begin{problems}
\item
Run \autoref{fig:incfixed} using (i) \texttt{synch} and (ii) \texttt{synchS}.
Report how many states were explored by Harmony for each module.
\item \label{ex:xy} \autoref{fig:xy} shows a Harmony program with two variables $x$
(initially 0) and $y$ (initially 100) that can be accessed through methods
\texttt{setX} and \texttt{getXY}.  An application invariant is that \texttt{getXY}
should return a pair that sums to 100.  Add the necessary synchronization code.
\item \label{ex:trylock} Implement \texttt{tryAcquire}(\textit{b}) as an additional
interface for both the \texttt{synch} and \texttt{synchS} modules.
This interface is like \texttt{acquire}(\textit{b}) but never blocks.  It
returns \texttt{True} if the binary semaphore was available (and now acquired) or \texttt{False}
if the binary semaphore was already acquired.
Hint: you do not have to change the existing code.
\item People who use an ATM often first check their balance and then withdraw
a certain amount of money not exceeding their balance.  A negative balance
is not allowed.  \autoref{fig:atm} shows two operations on bank accounts:
one to check the balance and one to withdraw money.
Note that all operations on accounts are carefully protected by a lock
(i.e., there are no data races).
The \texttt{customer}
method models going to a particular ATM and withdrawing money not exceeding
the balance.
Running the code through Harmony reveals that there is a bug.
It is a common type of concurrency bug known as \emph{Time Of Check Time Of
Execution} (TOCTOE).
\index{Time Of Check Time Of Execution}
\index{TOCTOE}
In this case, by the time the withdraw operation is performed,
the balance can have changed.
Fix the code in \autoref{fig:atm}.  Note, you should leave the
customer code the same. You are only allowed to change the
\texttt{atm\_} methods, and you cannot use the \textbf{atomic}
statement.
\end{problems}

@@

We will first implement the queue as a linked list.
The implementation in \autoref{fig:queue}
uses the \texttt{alloc} module for dynamic allocation
\index{dynamic allocation}%
of nodes in the list using \texttt{malloc}() and \texttt{free}().
\texttt{malloc}($v$) returns a new memory location initialized to $v$,
which should be released with \texttt{free}() when it is no longer in use.
The queue maintains a \texttt{head} pointer to the first element in the list
and a \texttt{tail} pointer to the last element in the list.
The \texttt{head} pointer is \texttt{None} if and only if the queue is empty.
(\texttt{None} is a special address value that is not the address of any
memory location.)

\texttt{queue.Queue()} returns a new queue object consisting of a \texttt{None} head
and tail pointer and a lock.
$\mathtt{queue.put}(q, v)$ and $\mathtt{queue.get}(q)$ both take a pointer $q$ to the
queue object because both may modify the queue.
Before they access the value of the head or tail of the queue they first obtain
the lock.
When they are done, they release the lock.

The implementation in \autoref{fig:queueMS}~\cite{MS98} is similar but uses
separate locks for the head and tail, allowing a \texttt{put} and a \texttt{get} operation
to proceed concurrently.  To avoid contention between the head and the tail,
the queue uses a dummy node at the head of the linked list.
Except initially, the dummy node is the last node that was dequeued.
Note that neither the \texttt{head} nor \texttt{tail} pointer are
ever \texttt{None}.
